---
title: Bayesian changepoint modeling using Nimble
author: Magnus Nygård Osnes
date: '2022-11-17'
slug: []
categories:
  - TLs
  - LineageHomology
  - Changepoint modeling
tags: []
image: ~
description: ~
toc: ~
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, collapse=T, cache=T, 'styler', tidy.opts=list(strict=T))
```

```{r loading files needed for script, echo=FALSE, include=FALSE}
#   ____________________________________________________________________________
#   Analysis                                                                ####


library(ape)
library(LineageHomology)
library(skygrowth)
library(grid)
library(scales) #Make more axis ticks.
library(gridExtra)
library(ggplot2)
library(phylodyn)
library(coefplot)

#Notes:
#Had to install an older version of RcppArmadillo to get the script running
#"install.packages("https://cran.r-project.org/src/contrib/Archive/RcppArmadillo/RcppArmadillo_0.9.900.3.0.tar.gz", repos=NULL, type="source")"

#Setup paths
path_main = "~/Dropbox/10000_Lovers/Phylogeography"
path_data = "~/Dropbox/10000_Lovers/01_LineageHomology/11_saved_data/"
path_results = "~/Dropbox/10000_Lovers/01_LineageHomology/12_saved_results/"
path_final_figures = "~/Dropbox/10000_Lovers/01_LineageHomology/99_Figures_for_paper/"

source("~/Dropbox/10000_Lovers/01_LineageHomology/sourcefunctions/environment_variables.R")
load(file=paste0(path_data,"Result_EUR.Rdata"))
Result_EUR=LineageHomology::reorder_LH(Result_EUR)
#RQ: Does meta-data explain the growth-rate of the biggest clades in period 2015-2018?
#1: Run Skygrowth on each TL.
#2: Extract clade wise growth-rate.
#3: Define explanatory variables for each clade: 
#4: Regress the growth rates on the metadata.


#   ____________________________________________________________________________
#   Set up treeclusters: Focus on the 20 largest from each location         ####

```

```{r script specific functions, echo=FALSE, include=FALSE}
clusters_and_times = function(result,main_tree,larger_than=10) {
  
  tree_clusters = list()
  time_shift_list = list() #For keeping track of time when shifting clusters. 
  for (i in 1:(which(result$Lineage_sizes<11)[1]-1))  #Loop over all lineages larger than 
  {
    cluster_tips = result$Taxa_names[i][[1]]
    tips_to_drop = main_tree$tip.label[main_tree$tip.label%in%cluster_tips==FALSE] #Tips to drop in this.
    mrca_node = getMRCA(main_tree,tip=cluster_tips)
    subtree = drop.tip(main_tree,tip =tips_to_drop)
    tree_clusters[[i]]= subtree
    time_shift_list[[i]] = nodeheight(main_tree, node=mrca_node)
  }
  return(list("clusters"=tree_clusters, "time_shifts"=time_shift_list))
}

neg_sizes = function(cluster_times) {
  neg_size_clusters = list()
  
  for(i in 1:length(cluster_times$time_shifts)) {
    tree_temp = cluster_times$clusters[[i]]
    class(tree_temp) = "phylo"
    #Estimate 1 Ne for every 1 year
    n_intervals = floor(max(nodeHeights(tree_temp))/1)
    neg_size_clusters[[i]]=skygrowth.map(tree_temp, res=n_intervals, maxiter=400)
  }
  neg_size_clusters
}

custom_skygrowth_plot <- function( fit,shift_time=NA,ggplot=TRUE, logy=TRUE,xname=NA,xlimits=NA){
  ne <- fit$ne
  fit$time<-fit$time+shift_time
  pldf <- data.frame( t = fit$time,  nemed = ne, nelb = fit$ne_ci[,1], neub = fit$ne_ci[,3] )
  pl <- ggplot2::ggplot( pldf, ggplot2::aes_( x = ~ t, y = ~ nemed)) + ggplot2::geom_line()+ ggplot2::ylab('Effective population size') + ggplot2::xlab(xname)
  pl <- pl + ggplot2::geom_ribbon( ggplot2::aes_( ymin = ~ nelb, ymax = ~ neub), fill = 'blue', alpha = .2)
  if(logy==T){
    pl <- pl + ggplot2::scale_y_log10()  
  }
  if(is.na(xlimits)==F){
    ind_select = which(fit$time>=xlimits)
    fit$ne_ci[ind_select,]; ylims = range(fit$ne_ci[ind_select,])
    pl<-pl+lims(x=c(xlimits,NA),y=ylims)
  }
  return(pl)
}

skygrowth_multiplot = function(neg_sizes,root_date,time_shift_list,xlimits, logy) {
  c1 = lapply(1:length(neg_sizes), FUN = function(x) custom_skygrowth_plot(neg_sizes[[x]],shift_time=(root_date+time_shift_list[[x]]), xname=paste("Cluster: ",x),xlimits=xlimits, logy=logy))
  n <- length(c1)
  nCol <- floor(sqrt(n))
  do.call("grid.arrange", c(c1, ncol=nCol))
}


return_growth_interval=function(clusterwise_output, period=c(2000,2017), shift_time=1750,time_shift_list) {
  ncluster = length(clusterwise_output)
  average_growth_rate = c()
  for(i in 1:ncluster){
    times = clusterwise_output[[i]]$time+shift_time+time_shift_list$time_shifts[[i]] #Adding Root date + cluster tmrca.
    inds = which(times>=period[1]&times<=period[2]) #indexes in time periods.
    growth_rate = mean(clusterwise_output[[i]]$growthrate[inds])
    average_growth_rate=c(average_growth_rate,growth_rate)
  } 
  return(average_growth_rate)
}


find_lineage_age = function(result, larger_than) {
  rel_inds = which(result$Lineage_sizes>larger_than)
  lineage_age = c()
  for(i in rel_inds){
    result$Taxa_names[[i]]
    youngest_sequence = max(name_date$dates[unlist(lapply(result$Taxa_names[[i]], FUN=function(x) which(x==name_date$name)))])
    tl_age = youngest_sequence-result$Halfedge_over_tmrca[i]
    lineage_age = c(lineage_age,tl_age)
  }
  lineage_age
}

define_explanatory_variables = function(result, larger_than) {
  #Sex distribution
  sex = sex_distributions(RLH = result,larger_than=larger_than)
  sex_variable = c()
  for(i in seq(1,nrow(sex),3)){ #Removes no data column.
    fem = sex$Count[i]
    mal = sex$Count[i+1]
    tmp=(mal)/sum(fem+mal)
    if(sum(fem+mal)<floor(larger_than/2)) { #If there are less than half of the minimum size we have nearly no information -> set Sex to NA. 
      tmp = NA  
    }
    sex_variable = c(sex_variable,tmp)
    
  }
  #MTR mosaic category: sets up a categorical variable.
  mtrDs = mtrD(RLH = result,larger_than=larger_than)
  mtr_variable = c()
  lineage_size = c() #defining this here for convenience.
  for(i in seq(1,nrow(mtrDs),7)){
    mtrDcounts = mtrDs$Count[i:(i+5)]
    maxind=which.max(mtrDcounts)
    mtr_variable=c(mtr_variable,mtrDs$State[i+(maxind-1)])
    lineage_size=c(lineage_size,mtrDs$Lineage_size[i+(maxind-1)]) #No reason to make it this complex (copypasted). 
  }
  
  #PenA state category
  penAs = penA_mosaic_distribution(RLH = result,larger_than=larger_than)
  penA_variable=c()
  for(i in seq(1,nrow(penAs),4)){
    penAscounts = penAs$Count[i:(i+3)]
    maxind=which.max(penAscounts)
    penA_variable=c(penA_variable,penAs$State[i+(maxind-1)])
  }
  lineage_age = find_lineage_age(result=result, larger_than=larger_than)
  d1 = data.frame(lineage_size,penA_variable,mtr_variable,sex_variable, lineage_age)
  return(d1)
}


```

```{r Analysis and environment, echo=FALSE, include=FALSE}
#Results for Norway
cluster_times_NOR = clusters_and_times(Result_NOR, main_tree = tree)
# neg_sizes_NOR = neg_sizes(cluster_times_NOR)
# save(neg_sizes_NOR,file=paste0(path_results,"neg_NOR.Rdata"))
load(file=paste0(path_results,"neg_NOR.Rdata"))
# pdf(paste0(path_results,"neg_NOR_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_NOR,root_date=1750, time_shift_list=cluster_times_NOR$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_NOR_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_NOR,root_date=1750, time_shift_list=cluster_times_NOR$time_shifts,xlimits=NA, logy=F)
# dev.off()


#Results for AUSTRALIA
cluster_times_AUS = clusters_and_times(Result_AUS, main_tree = tree)
# neg_sizes_AUS = neg_sizes(cluster_times_AUS)
# save(neg_sizes_AUS,file=paste0(path_results,"neg_AUS.Rdata"))
load(file=paste0(path_results,"neg_AUS.Rdata"))
# pdf(paste0(path_results,"neg_AUS_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_AUS,root_date=1750, time_shift_list=cluster_times_AUS$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_AUS_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_AUS,root_date=1750, time_shift_list=cluster_times_AUS$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Results for USA
cluster_times_USA = clusters_and_times(Result_USA, main_tree = tree)
# neg_sizes_USA = neg_sizes(cluster_times_USA)
# save(neg_sizes_USA,file=paste0(path_results,"neg_USA.Rdata"))
load(file=paste0(path_results,"neg_USA.Rdata"))
# pdf(paste0(path_results,"neg_USA_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_USA,root_date=1750, time_shift_list=cluster_times_USA$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_USA_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_USA,root_date=1750, time_shift_list=cluster_times_USA$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Results for Europe
cluster_times_EUR = clusters_and_times(Result_EUR, main_tree = tree)
# neg_sizes_EUR = neg_sizes(cluster_times_EUR)
# save(neg_sizes_EUR,file=paste0(path_results,"neg_EUR.Rdata"))
load(file=paste0(path_results,"neg_EUR.Rdata"))
# pdf(paste0(path_results,"neg_USA_top20.pdf"),width=15,height=15)
#skygrowth_multiplot(neg_sizes_EUR,root_date=1750, time_shift_list=cluster_times_EUR$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_USA_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_USA,root_date=1750, time_shift_list=cluster_times_USA$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Notes: Sampling is different between the locations
#The Australian collection, sampling stops in 2017 and starts in 2016 + some few genomes in 2004
#For USA there are a few samples from 1997, some from 2009, 2010, and the main bulk from 2016-2018, 2017 has much less than the neighboring years. 
#Thus we might extract the average growth rate in the period 2015-2017 for Australia, 2013-2018 for Norway, and  2009-2019 for USA? 


#   ___________________________________________________________________________________________
#   Can the growth rate be explained by explanatory variables such as sex distribution       ####

#Extract the growth rates on the intervals 2014-2017: This may not be meaningful. 
r_NOR = return_growth_interval(clusterwise_output=neg_sizes_NOR, time_shift_list=cluster_times_NOR)
r_AUS = return_growth_interval(clusterwise_output=neg_sizes_AUS, time_shift_list=cluster_times_AUS)
r_USA = return_growth_interval(clusterwise_output=neg_sizes_USA, time_shift_list=cluster_times_USA)
r_EUR = return_growth_interval(clusterwise_output=neg_sizes_EUR, time_shift_list=cluster_times_EUR)
r_all = c(r_NOR,r_AUS,r_USA,r_EUR)

d1 = define_explanatory_variables(Result_NOR,larger_than=10)
d2 = define_explanatory_variables(Result_AUS,larger_than=10)
d3 = define_explanatory_variables(Result_USA,larger_than=10)
d4 = define_explanatory_variables(Result_EUR,larger_than=10)


#Gather data
total_data = rbind(d1,d2,d3,d4)
locations = c(rep("Norway",length(r_NOR)),rep("Australia",length(r_AUS)),rep("USA",length(r_USA)), rep("EUR",length(r_EUR)))
total_data = cbind(r_all,locations,total_data) #Add growth rate
colnames(total_data)= c("GrowthRate","locations", "tlSize","penA","mtr", "sexDistribution", "lineage_age") 

#   ____________________________________________________________________________
#   Report starts from here                                                 ####


```

## Modeling the growth rate of transmission lineages 

We are working on a project where we try to illuminate some aspect of lineage dynamics of *Neisseria gonorrhoeae*.
We are focusing on four locations divided into small scale: Norway and Australia, and large scale: Europe and the USA. For each of the four geographical locations we have used [LineageHomology](https://github.com/magnusnosnes/LineageHomology) to define transmission lineages wrt. each of those locations. A "transmission lineage" is group of tips on a phylogeny that are  are estimated to stem from the same importation event. Hence such groups may be used to retrospectively assess how important a single importation events in terms of onward transmission of the disease. 
## **Research question** 
In this post I look at the relation between the growth rate of the lineages in the observed gender distrbution of the infected hosts in the transmission lineages. 
The reason we wanted to investigate this question is that the the observed gender distrbution of the hosts likely reflects the sexual network structure (such as hetersexual communities, men who have sex with men, or mixed). And we wanted to see if there were differences in the growth rate of the lineages depending on the networks they "inhabit".

## A look at the data

```{r, echo=FALSE, include = T, warning=F, message=F}
g1 = ggplot(total_data, aes(x=sexDistribution, y=GrowthRate, color=locations, fill=locations))+theme_bw()+
  geom_point()+ylab("Growth rate")+xlab("Gender distrbution in TL - fraction men")+theme(legend.position =c(0.2,0.8))+
  ylim(c(0,5))
g1

```

\

Looking at the data we thought that there may be some threshold in the observed fraction of men that are indicative of the sexual network structure. Instead of trying to define some threshold such as 70% > indicates MSM, we wanted to define a changepoint model where this threshold is estimated along with the differences in growth rate. 

<!--  ### Size of the transmission lineages

```{r, echo=FALSE, include = T, warning=F, message=F, }
g1 = ggplot(total_data,aes(x=sexDistribution, y=tlSize, color=locations, fill=locations))+theme_bw()+geom_point()+ylab("Transmission lineage size")+theme(legend.position="none")
#+geom_smooth(method="lm", alpha=0.15)
```
 
```{r, echo=FALSE, include = T, warning=F, message=F, fig.width=8,fig.height=8}
g2 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=locations, y=tlSize,color=locations))+theme(legend.position = "none")+ylab("Transmission lineage size")
g3 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=mtr, y=tlSize))+ylab("Transmission lineage size")
g4 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=penA, y=tlSize))+ylab("Transmission lineage size")
grid.arrange(g2,g3,g4,nrow=3,ncol=1)

#boxplot(total_data$GrowthRate~as.factor(total_data$locations),ylab="Growth Rate", xlab="Transmission lineage size")
```
--> 

```{r, echo=FALSE, include = F}
total_data_modified = total_data #Modfied version of data
total_data_modified$penA[which(total_data_modified$penA=="nd")]=NA #set nd to NA
total_data_modified$penA[which(total_data_modified$penA=="semi-mosaic")]="mosaic" #set nd to NA
total_data_modified$mtr[which(total_data_modified$mtr=="nd")]=NA
total_data_modified$mtr[total_data_modified$mtr!="non-mosaic"]="mosaic"
total_data_modified$mtr[total_data_modified$mtr=="semi-mosaic"]="mosaic"



```



```{r, echo=FALSE, include = T}
#Pairwise comparisons between Norway and Australia and Europe and USA
total_data_modified = total_data_modified[complete.cases(total_data_modified),]
total_data_small_scale = total_data_modified[which(total_data_modified[,"locations"] %in% c("Norway", "Australia")),]
total_data_large_scale = total_data_modified[which(total_data_modified[,"locations"] %in% c("EUR", "USA")),]


small_scale = total_data_small_scale
small_scale$locations = factor(small_scale$locations)
#Refactor to change levels

large_scale = total_data_large_scale
large_scale$locations = factor(large_scale$locations)
## NB NEED TO FACTOR COUNTRES differently when working with the array this way. 
```


## Bayesian changepoint modeling

To describe the threshold we specify an threshold "alpha", such that the mean growth rate in is different before and after . 
We use the r-package [NIMBLE](https://r-nimble.org/) for statistical programming.


## Changepoint model

### Defining data for NIMBLE

Our data looks like:

```{r, echo=T, include = T, warning=F, message=F, eval=T}
library(nimble)

head(small_scale[,c("GrowthRate","sexDistribution", "locations")])
summary(small_scale[,c("GrowthRate","sexDistribution", "locations")])
```
 
NIMBLE requires the data defined in a named list with of constants, in addition to other variables that will be used in the nimble model: such as the length of the data "n" and 
 
 
```{r, echo=T, include = T, warning=F, message=F, eval=T}
data <- list(y=small_scale[,"GrowthRate"]) #The growthrate that we are trying to explain.
n = length(small_scale[,"GrowthRate"]) #Number of data points
constants = list(n=n, genderDistribution=small_scale$sexDistribution, #Constants to be used in the Nimblemodel
                 location=as.numeric(as.factor(small_scale$locations)))

```

### Defining the changepoint model in NIMBLE

Next we define the NIMBLE-model for an unobserved changepoint in the fraction of men indicating a different sexual network structure. We assume the the errors follow a normal distribution around a mean that is different above and below the threshold. 
 
```{r, echo=T, include = T, warning=F, message=F, eval=T}
changepoint <- nimbleCode({
  alpha ~ dunif(0.7,1) #Changepoint
  sigma1 ~ dunif(0, 5)        # prior for variance components based on Gelman (2006)
  for(k in 1:2){
    mu1[k]~ dnorm(0, sd = 10) 
    mu2[k]~ dnorm(0, sd = 10) 
  }
  
  for(i in 1:n) {
    turn_on[i] <- step(genderDistribution[i]-alpha) #0 if y[i] is smaller than and 1 otherwise.
    mug[i] <- mu1[location[i]]*turn_on[i]+mu2[location[i]]*(1-turn_on[i]) #Switch means depending on if we are above or below the changepoint.
    
    #Likelihood of the data (growthrate)
    y[i] ~ dnorm(mug[i],sigma1)
  }
})

```

### Setting initial values

Before running we have to set the initial values of the parameteres for the MCMC.
 
```{r, echo=T, include = T, warning=F, message=F, eval=T}
inits <- list(mu1 =c(1,1),mu2=c(1,1),
              sigma1 = 1,
              #sigma2 = 1,
              alpha=0.8)
```

### Running the changepoint model 

We run the model on the small scale data from Australia and Norway " and summarize the results on 
 
```{r, echo=T, include = T, warning=F, message=F,cache=T, eval = T}
#Run the MCMC
samples_small_scale = nimbleMCMC(code = changepoint, constants = constants,
                       data = data, inits = inits,
                       nchains = 2, niter = 1000000,
                       summary = TRUE, WAIC = TRUE,thin = 100,
                       monitors = c('mu1','mu2','sigma1',
                                    #'sigma2',
                                    'alpha'))

```

Next, we summarize results using the MCMCvis package check that mixing is okay using the effective sample size (last column)

### Australia and Norway

```{r, echo=T, include = T, warning=F, message=F, eval=T}
library(MCMCvis)
MCMCsummary(object = samples_small_scale$samples)
MCMCplot(object = samples_small_scale$samples,params=c("mu1", "mu2"))
#MCMCtrace(object = samples_small_scale$samples,params=c("alpha")) #Look at the traceplots
```

### Europe and the USA
 
```{r, echo=F, include = F, warning=F, message=F, eval=T}
data <- list(y=large_scale[,1])
constants = list(n=n, genderDistribution=large_scale$sexDistribution, location=as.numeric(as.factor(large_scale$locations)))
inits <- list(mu1 =c(1,1),mu2=c(1,1),
              sigma1 = 1,
              #sigma2 = 1,
              alpha=0.8)
#Changepoint 1
samples_large_scale = nimbleMCMC(code = changepoint, constants = constants,
                       data = data, inits = inits,
                       nchains = 2, niter = 1000000,nburnin = 1000000/2,
                       summary = TRUE, WAIC = TRUE,thin = 100,
                       monitors = c('mu1','mu2','sigma1',
                                    #'sigma2',
                                    'alpha'))
```
 
```{r, echo=T, include = T, warning=F, message=F, eval=T}
MCMCsummary(object = samples_large_scale$samples)
MCMCplot(object = samples_large_scale$samples,params=c("mu1", "mu2"))
#MCMCtrace(object = samples_large_scale$samples,params=c("alpha")) #Look at the traceplots
```


## Regression model

We use the same dataset but change to a regression model. 

## Defining the model

```{r, echo=T, include = T, warning=F, message=F, eval=T}
regression <- nimbleCode({
  
  sigma1 ~ dunif(0, 5)        # prior for variance components based on Gelman (2006)
  
  #Priors for regression slope
  for(k in 1:2){
    mu1[k]~ dnorm(0, sd = 10) 
    beta[k] ~ dnorm(0, sd = 10) #Regression slope
  }
  
  
  
  for(i in 1:n) {
    mug[i] <- mu1[location[i]]+beta[location[i]]*genderDistribution[i] #Define the expected value
    
    #Likelihood of the growth rate
    y[i] ~ dnorm(mug[i],sigma1)
  }
})


```

### Norway and Australia
 
```{r, echo=T, include = T, warning=F, message=F, eval=T}
data <- list(y=small_scale[,1])
constants = list(n=n, genderDistribution=small_scale$sexDistribution,
                 location=as.numeric(as.factor(small_scale$locations)))
inits <- list(mu1 =c(1,1),
              sigma1 = 1,
              #sigma2 = 1,
              beta=c(0.8,0.8))

regression_samples_small_scale = nimbleMCMC(code = regression, constants = constants,
                       data = data, inits = inits,
                       nchains = 2, niter = 1000000,nburnin = 1000000/2,
                       summary = TRUE, WAIC = TRUE,thin = 100,
                       monitors = c('mu1',
                                    'sigma1',
                                    'beta'))
par(mfrow=c(1,1))
MCMCsummary(object = regression_samples_small_scale$samples)
MCMCplot(object = regression_samples_small_scale$samples,params=c("mu1","beta"))
#MCMCtrace(object = regression_samples_small_scale$samples,params=c("beta"))

```
 
### Europe and the USA 

```{r, echo=T, include = T, warning=F, message=F, eval=T}
data <- list(y=large_scale[,1])
constants = list(n=n, genderDistribution=large_scale$sexDistribution,
                 location=as.numeric(as.factor(large_scale$locations)))
inits <- list(mu1 =c(1,1),
              sigma1 = 1,
              #sigma2 = 1,
              beta=c(0.8,0.8))

regression_samples_large_scale = nimbleMCMC(code = regression, constants = constants,
                       data = data, inits = inits,
                       nchains = 2, niter = 1000000,nburnin = 1000000/2,
                       summary = TRUE, WAIC = TRUE,thin = 100,
                       monitors = c('mu1',
                                    'sigma1',
                                    'beta'))
par(mfrow=c(1,1))
MCMCsummary(object = regression_samples_large_scale$samples)
MCMCplot(object = regression_samples_large_scale$samples,params=c("mu1","beta"))
#MCMCtrace(object = regression_samples_large_scale$samples,params=c("beta"))
```

